---
title: "Multigroup multilevel models"
author: "(Marcello Gallucci)"
nickname: multiall
topic: general
category: example
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE,results='hide'}
source("../R/constants.R")
source("../R/functions.R")

```


`r keywords("SEM, multilevel, multigroup, lavaan, SEMLj, jamovi")`

`r version("0.5.0")`
`r draft`
`r tobecontinued`

In this example we show how to estimate a multilevel, multigroup path analysis using `r modulename`. 

We show input of `r modulename` _syntax sub-module_, because multilevel models can be run in `r modulename` only in the syntax sub-module.

Much of the output that `r modulename` produces is labeled as it is in lavaan R package output, so lots of information can be found in [lavaan help and tutorials](https://lavaan.ugent.be/tutorial/tutorial.pdf).

Obviously, one can run the models discussed here directly in R using `lavaan` package, if `r jamovi` is not the software one is using at the moment.

# The data

For this example we are going to use the `Demo.twolevel` dataset present in `lavaan` R package, which is useful for estimating a multilevel model, with a small modification. The dataset contains 200 clusters, within each 6 items (y1-y2) and 3 continuous variables (x1-x3) are measured. Across clusters (between level) two covariates (w1 and w2) are measured . 

We now imagine that the 200 clusters are evenly divided in two groups, `g1` and `g2`. Had this dataset been based on a survey in 200 school classes, for instance, one can imagine that 100 classes came from one state (`g1`) and the other 100 from another state.  We created the `group` variable on the spot for the sake of this exercise.

`r pic("pics/example4/input_data1.png")`


We want to estimate the multilevel model (cf [lavaan help page](https://lavaan.ugent.be/tutorial/multilevel.html)), in the whole sample, then separated for the two groups, and test some differences in parameters between the two groups.


# The model (whole sample)

The model  has a latent variable measured within clusters (`fw`) with 3 items `y1,y2,y3`. At the within-cluster level, the latent variable is predicted by the three covariates `x1`, `x2` and `x3`.

At the between-cluster level, the same latent variable is present measured with the same items (`fb=y1+y2+y3`), but it is predicted by the two between-cluster covariates (`w1` and `w2`). Here is the model for the whole sample. 

```{r, eval=FALSE}

    level: within
        fw =~ y1 + y2 + y3
        fw ~ x1 + x2 + x3
    level: between
        fb =~ y1 + y2 + y3
        fb ~ w1 + w2


```



# Output

As soon as we set the input, we obtain the whole set of results tables. We do not go through the results now, because this is not the model we are focusing on, but we need to be sure that the estimation converges in the whole sample, so we are ready for the multigroup analysis. In fact, if the model does not converge in the whole sample, it is difficult to make it converge in multigroup analysis.

`r pic("pics/example4/output_info1.png")`


# Multigroup model

We now set the multigroup analysis on top of the multilevel model. The model syntax needs to be repeated for the two groups explicitly. **Important point**: whereas standard multigroup analysis only requires to define a grouping variable in the `Multi-group analysis` panel, here we need to specify a grouping variable in the `Multigroup analysis` panel and the grouping structure in the syntax. 

There you go: First, we set the multigroup variable

`r pic("pics/example4/input_group.png")`

and then we issue the following syntax:

`r pic("pics/example4/input_syntax2.png")`

`r tobecontinued`





# Related examples
`r include_examples()`

`r issues()`